Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-18 sections/01-introduction.tex
Update on Overleaf.

In contrast to conventional forecasting methods, nowcasting methods focus on predictions for the present, the immediate future, and the recent past \parencite{banbura2013now} and are now widely used in fields such as economics and medicine \parencite{bok2018macroeconomic, Wolffram2023}.
Nowcasting has its origins in meteorology, and the methods were initially developed to describe the current state of the weather in detail and to predict the expected change on a time scale of a few hours \parencite{browning1989nowcasting,schmid2019nowcasting}. 
In economics, nowcasting is used to predict statistics on the current economic situation, for example, the gross domestic product, which is collected with low frequency and is available with a considerable time delay \parencite{banbura2013now}.
Nowcasting methods use high-frequency indicators related to the target variable.
Thus, the nowcasts can produce early and ongoing estimates for the target variable during the relevant period \parencite{castle2017forecasting}. 
In medicine, epidemic nowcasting assesses the current situation during an ongoing epidemic, considering the main pathogenic, epidemiological, clinical, and socio-behavioral factors \parencite{wu2021nowcasting}. 
For example, the nowcasting method can correct the daily COVID case numbers for events that have occurred but have not yet been reported \parencite{gunther2021nowcasting}. 
In the medical context, nowcasting is an estimation procedure in which the value of a target variable for a specific time is estimated based on current preliminary measurements, which are only finalized later. 
At the time of nowcasting, estimates of the target variable can be made for the current time and for a past time; the predicted change in the target variable is then the difference between the two nowcasts. 
A nowcast is consistent with the evolution of the target variable if the difference between the actual values at the current time and the actual values at a past time has the same sign as the predicted change. 
Like forecasts, nowcasts are often evaluated and compared in the literature based on performance measures such as the \ac{rmse} \parencite{gunther2021nowcasting} or probabilistic measures \parencite{Wolffram2023}. 
The aspect of trending ability is not considered in the literature to the best of our knowledge. 
However, trending evaluation reveals a method's capability of predicting the development of the target variable, for example, the number of cases of an epidemic, that can be the foundation of decisions on the introduction or cancellation of policy measures. \todo{Satz ist noch nicht gut}

As outlined above, trending evaluation is crucial for measurements, predictions, and forecasts. 
However, regardless of the application area, most conventional method comparisons and assessments consider local absolute deviations without assessing whether the correct direction of change is being predicted or measured. In this paper, we want to complement the current evaluation measures with trending measurements and propose evaluation methods that assess the trending ability of a method.

The main contributions of this paper are summarized below
\begin{itemize}
\item We formalize trending and show how the predicted changes can be computed to assess trending for practical relevance. 
\item We present different variants of trending measures that consider either noiseless data or data with noise and small non-informative changes.
\item We introduce the conditional trending plot, a new graphical method for assessing local trending behavior.
\item We introduce bootstrap methods for calculating confidence intervals for the presented treding ratios.
\item We extend the concept of trending to probabilistic predictions.
\item We apply trending evaluation to three applications: Measurement, nowcasting, and forecasting data. 
\item We provide a ready-to-use code for trending evaluation. The code and the source code to replicate the results of our study are available in our public repository at ... .
\end{itemize}

The remainder of this paper is structured as follows. 
In Section \ref{sec:trending}, we formalize the concept of trending independently of the application domain and investigate quantification approaches for noise-free and noisy data. Noisy data refers to data where noise or unsystematic effects are present in the true values or predictions.
In particular, we introduce a new graphical method for evaluating trending, the conditional trending plot, which can represent the simultaneous evaluation of different intervals and asymmetries. 
Bootstrap methods for calculating confidence intervals for trending measures are also presented. 
In Section \ref{sec:application}, we show the results of applying our method to several practical examples from measurement, forecasting, and nowcasting. 
We conclude our paper in Section~\ref{sec:conclusion}.

>>>>

In contrast to conventional forecasting methods, nowcasting methods focus on predictions for the present, the immediate future, and the recent past \parencite{banbura2013now} and are now widely used in fields such as economics and medicine \parencite{bok2018macroeconomic, Wolffram2023}.
Nowcasting has its origins in meteorology, and the methods were initially developed to describe the current state of the weather in detail and to predict the expected change on a time scale of a few hours \parencite{browning1989nowcasting,schmid2019nowcasting}. 
In economics, nowcasting is used to predict statistics on the current economic situation, for example, the gross domestic product, which is collected with low frequency and is available with a considerable time delay \parencite{banbura2013now}.
Nowcasting methods use high-frequency indicators related to the target variable.
Thus, the nowcasts can produce early and ongoing estimates for the target variable during the relevant period \parencite{castle2017forecasting}. 
In medicine, epidemic nowcasting assesses the current situation during an ongoing epidemic, considering the main pathogenic, epidemiological, clinical, and socio-behavioral factors \parencite{wu2021nowcasting}. 
For example, the nowcasting method can correct the daily COVID case numbers for events that have occurred but have not yet been reported \parencite{gunther2021nowcasting}. 
In the medical context, nowcasting is an estimation procedure in which the value of a target variable for a specific time is estimated based on current preliminary measurements, which are only finalized later. 
Like forecasts, nowcasts are often evaluated and compared in the literature based on performance measures such as the \ac{rmse} \parencite{gunther2021nowcasting} or probabilistic measures \parencite{Wolffram2023}. 
The aspect of trending ability is not considered in the literature to the best of our knowledge. 
However, trending evaluation reveals a method's capability of predicting the development of the target variable, for example, the number of cases of an epidemic, that can be the foundation of decisions on introducing or canceling policy measures. \todo{Satz ist noch nicht gut}

As outlined above, trending evaluation is crucial for measurements, predictions, and forecasts. 
However, regardless of the application area, most conventional method comparisons and assessments consider local absolute deviations without assessing whether the correct direction of change is being predicted or measured. 
In this paper, we develop trending evaluation methods that complement current evaluation measures with trending measurements.

The main contributions of this paper are manifold.
\begin{itemize}
\item We formalize trending and present different variants of trending measures that consider either noiseless data or data with noise and small non-informative changes.
\item We introduce the conditional trending plot, a new graphical method for assessing local trending behavior, and review bootstrap methods for calculating confidence intervals.
\item We extend the concept of trending to probabilistic predictions.
\item We apply trending evaluation to three applications: Measurement, nowcasting, and forecasting data. 
\item We provide a ready-to-use code for trending evaluation. The code and the source code to replicate the results of our study are available at ... .
\end{itemize}

The remainder of this paper is structured as follows. 
Section \ref{sec:trending} formalizes trending and investigates several extensions, such as noise-aware methods, confidence intervals, and probabilistic forecasts and nowcasts. 
In particular, we introduce a new graphical method for evaluating trending, the conditional trending plot, which can represent the simultaneous evaluation of different intervals and asymmetries. 
In Section \ref{sec:application}, we show the results of applying our method to several practical examples from measurement, forecasting, and nowcasting. 
We conclude the paper in Section~\ref{sec:conclusion}.

### 2024-03-18 sections/03-trending.tex
Update on Overleaf.

As discussed in Section~\ref{sec:introduction}, the term \enquote{trending} denotes the alignment of a detected true development with the development indicated by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and define different methods of measuring and visualizing trending.
Development always has respect to a time horizon, that is, the amount of time passed. 
In general, trending can be assessed for different horizons $l$ separately, and the considered horizons' relevance must be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal} whenever the application area is insignificant.
Before introducing measures to evaluate trending, we remark on calculating the predicted change and notation in Section~\ref{subsec:notation}.
Section~\ref{subsec:trending-four-quadrant-plot} introduces trending basics and visualizes trending in a synthetic example.
In Section~\ref{subsec:trending-measures}, various methods from other disciplines and tasks are applied and examined.
The approaches are extended to account for noise and non-systematic effects in Section~\ref{subsec:trending-noise} and bootstrap methods for computing confidence intervals for the measures in the previous sections are briefly introduced.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method for local trending assessment.
Trending assessment for probabilistic nowcasts and forecasts is treated in Section~\ref{subsec:trending-probabilistic}.

@@

As noted above, trending refers to observed and predicted changes over a time horizon $l$.
The \textit{observed} change is straightforward to compute for all types of signals.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values for nowcasting or forecasting, or gold standard measurements up to time $T$.
The sequence of observed changes is then given by the differences of values in $\mathbf{y}$ with horizon $l$, that is
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t} - y_{t-l}) \quad \text{for}\ t = l, \dots, T.
\end{equation}

@@

The definition of the predicted change depends on the context.
Table~\ref{tab:notation} summarizes the notation for the different signal types and the computation of the predicted change.
For nowcasting, let $x_{t | \tau}$ denote the nowcast for time $t$ computed with the knowledge of time $\tau$. 
We call $t$ the \textit{forecast time} and $\tau$ the \textit{issue time}.
We compute the predicted change by
\begin{equation}
  \diffxl =
\begin{cases}
(x_{t|t} - x_{t-l|t})^T_{t=l} & \text{if } y_{t-l} \text{ is not known at time } t, \\
(x_{t|t} - y_{t-l})^T_{t=l} & \text{otherwise}.
\end{cases} \label{eq:diffxl_nowcasting}
\end{equation}
When computing the predicted change of a nowcast for a time $t$, we use the best knowledge available at that time $t$, and the true value might not be known yet. 
Thus, the predicted can be computed with the knowledge of time $t$, and the trending assessment can be used in practice at time $t$.

The notation is similar for forecasting: Let $x_{t | \tau}$ denote the forecast for forecast time $t$ and issue time $\tau$. 
The predicted change is computed by 
\begin{equation} \diffxl = (x_{t|t-l} - y_{t-l})^T_{t=l} \label{eq:diffxl_forecasting}
\end{equation}
The indices are chosen to be consistent with $\diffyl$. 
If the true value $y_{t-l}$ is not known at time $t-l$, a similar modification can be made as in Equation~\eqref{eq:diffxl_nowcasting}.

@@

The trending assessment is the same for all three application areas, given the notation for the respective application of Section~\ref{subsec:notation}.
In the following, we omit the difference horizon $l$ for ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common horizon $l$ as defined in Table~\ref{tab:notation}.
As noted above, a signal's trending is perfect if all the change directions are predicted correctly; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Consequently, when evaluating the trending, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot~\parencite[see, e.g., ][]{Saugel2015,perrino1998intraoperative}. 
In a four-quadrant plot, the occured changes are plotted with the predicted changes, that is, $(\diffyt, \diffxt)$ for $t = 1, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays the prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct trending for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.

@@

The four-quadrant plot is intuitive to interpret, and the magnitude and direction of change are shown simultaneously.
It can be extended by including information on the date in the point color.
In Figure~\ref{fig:trending_basic_4q_sample_color}, the point colors turn from blue to green for higher time indices $t$.
The horizon is always one time step.
However, four-quadrant plots become crowded for larger datasets, and sequential information on the differences is complex to assess thoroughly.
There are other visualization techniques, such as polar plots or the Bland-Altmann analysis.
However, they lack the four-quadrant plot's clarity and intuition without adding more information on trending~\parencite{Saugel2015}.

>>>>

As discussed in Section~\ref{sec:introduction}, the term \enquote{trending} denotes the alignment of a detected true development with the development indicated by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and define different methods of measuring and visualizing trending.
Development always has respect to a time horizon, that is, the amount of time that has passed. 
In general, trending can be assessed for different horizons $l$ separately, and the considered horizons' relevance must be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal} whenever the application area is insignificant.
Before introducing measures to evaluate trending, we remark on calculating the predicted change and notation in Section~\ref{subsec:notation}.
Section~\ref{subsec:trending-four-quadrant-plot} introduces trending basics and visualizes trending in a synthetic example.
In Section~\ref{subsec:trending-measures}, various methods from other disciplines and tasks are applied and examined.
The approaches are extended to account for noise and non-systematic effects in Section~\ref{subsec:trending-noise}, and bootstrap methods for computing confidence intervals for the measures in the previous sections are briefly introduced.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method for local trending assessment.
Trending assessment for probabilistic nowcasts and forecasts is treated in Section~\ref{subsec:trending-probabilistic}.

@@

As noted above, trending refers to observed and predicted changes over a time horizon $l$.
The \textit{observed} change is straightforward to compute for all types of signals.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values for nowcasting or forecasting, or gold standard measurements up to time $T$.
The sequence of observed changes is then given by the differences of values in $\mathbf{y}$ with horizon $l$, that is,
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t} - y_{t-l}) \quad \text{for}\ t = l, \dots, T.
\end{equation}

@@

The definition of the \textit{predicted} change depends on the context.
Table~\ref{tab:notation} summarizes the notation for the different signal types and the computation of the predicted change.
For nowcasting, let $x_{t | \tau}$ denote the nowcast for time $t$ computed with the knowledge of time $\tau$. 
We call $t$ the \textit{forecast time} and $\tau$ the \textit{issue time}.
We compute the predicted change by
\begin{equation}
  \diffxl =
\begin{cases}
(x_{t|t} - x_{t-l|t})^T_{t=l} & \text{if } y_{t-l} \text{ is not known at time } t, \\
(x_{t|t} - y_{t-l})^T_{t=l} & \text{otherwise}.
\end{cases} \label{eq:diffxl_nowcasting}
\end{equation}
When computing the predicted change of a nowcast for a time $t$, we use the best knowledge available at that time $t$, and the true value might not be known yet. 
Thus, the predicted change can be computed with the knowledge of time $t$, and the trending assessment can be used in practice at time $t$.

The notation is similar for forecasting: Let $x_{t | \tau}$ denote the forecast for forecast time $t$ and issue time $\tau$. 
The predicted change is computed by 
\begin{equation} \diffxl = (x_{t|t-l} - y_{t-l})^T_{t=l} \label{eq:diffxl_forecasting}
\end{equation}
The indices are chosen to be consistent with $\diffyl$. 
If the true value $y_{t-l}$ is not known at time $t-l$, a similar modification as in Equation~\eqref{eq:diffxl_nowcasting} can be made.

@@

The trending assessment is the same for all three application areas, given the notation for the respective application of Section~\ref{subsec:notation}.
In the following, we omit the difference horizon $l$ for ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common horizon $l$ as defined in Table~\ref{tab:notation}.
As noted above, a signal's trending is perfect if all the change directions are predicted correctly; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Consequently, when evaluating the trending, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot~\parencite[see, e.g., ][]{perrino1998intraoperative,Saugel2015}. 
In a four-quadrant plot, the occured changes are plotted with the predicted changes, that is, $(\diffyt, \diffxt)$ for $t = l, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays the prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct trending for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.

@@

The four-quadrant plot is intuitive to interpret, and the magnitude and direction of change are shown simultaneously.
It can be extended by including information on the date in the point color.
In Figure~\ref{fig:trending_basic_4q_sample_color}, the point colors turn from blue to green for higher time indices $t$, that is, $(\diffyt[l], \diffxt)$
However, four-quadrant plots become crowded for larger datasets, and sequential information on the differences is complex to assess thoroughly.

There are other visualization techniques, such as polar plots or the Bland-Altmann analysis.
However, they lack the four-quadrant plot's clarity and intuition without adding more information on trending~\parencite{Saugel2015}.
